Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-18 manuscript/manuscript.tex
fixed subsection numbering [ci skip]
addressing @brookehus' comments

@@

\subsection*{Hidden Markov state models}

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs require the dynamics in some hidden (unobserved) state space to be Markovian.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a (small) transition matrix between the $m$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.

An HMM estimation usually yields a small model where each hidden state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully.
Given that an HMM is a generalization of an MSM,
we can compute all physico-chemically relevant quantities available for MSMs, too.

@@

While there exist automated schemes~\cite{husic-optimized} for cross-val\-\-i\-dat\-ed optimization in the full hyper-parameter space,
we chose to adopt a sequential approach where only the hyper-pa\-ra\-me\-ters of the current stage are optimized.
This approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling choices.

@@

\begin{figure}
\includegraphics{figure_6}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
(a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however,
(b)~can be enhanced if the system is prepared in the non\-equi\-lib\-ri\-um condition $\mathcal{S}_1$.
The solid/orange lines denote the maximum likelihood MSM result;
the dashed/blue lines and the the shaded areas indicate sample means and~$95\%$ confidence intervals computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.}
\label{fig:msm-exp-obs}
\end{figure}

>>>>

\hyphenation{val-i-dat-ed}

@@

\subsection{Hidden Markov state models}

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs require the dynamics in some hidden (i.e., unobserved) state space to be Markovian.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a (small) transition matrix between the $m$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.

An HMM estimation usually yields a model with a small number of (hidden) states
where each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
Given that an HMM is a generalization of an MSM,
we can compute all physico-chemically relevant quantities available for MSMs, too.

@@

While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter space,
we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized.
This approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling choices.

@@

\begin{figure}
\includegraphics{figure_6}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
(a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however,
(b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.
The solid/orange lines denote the maximum likelihood MSM result;
the dashed/blue lines and the the shaded areas indicate sample means and~$95\%$ confidence intervals computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.}
\label{fig:msm-exp-obs}
\end{figure}
